Efficient ancestry and mutation simulation with msprime 1.0

### 2021-11-18 paper.tex
Trim down mutations section
Closes #203

As well as providing a new API
that emphasises the logical split between ancestry and mutation simulation,
we have greatly extended the sophistication of
\msprime's mutation generation engine for version 1.0,
achieving near feature-parity with \SeqGen.
We support a large number of mutation models, including the
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
Other models, such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation},
can be defined easily and efficiently in
user code by specifying a transition matrix between
any number of alleles (which can be arbitrary unicode strings).
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
\Pyvolve~\citep{spielman2015pyvolve}.

@@

Simulating mutations in \msprime\ is efficient.
Fig.~\ref{fig-mutations-perf} shows the time required to generate
mutations (using the default JC69 model) on
simulated tree sequences for a variety of mutation
rates as we vary the number of samples
(Fig.~\ref{fig-mutations-perf}A) and the sequence
length (Fig.~\ref{fig-mutations-perf}B).
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with 508,125 edges.
This efficiency for large numbers of trees is possible because
the tree sequence encoding allows us to generate mutations
on an edge-by-edge basis
(see Fig.~\ref{fig-ts-example} and the~\nameref{app-mutation-algorithm}
appendix),
rather than tree-by-tree and branch-by-branch as would otherwise be required.
In the above example from Fig.~\ref{fig-mutations-perf}B,
if we generated mutations tree-by-tree, we would have to iterate over 273,887,838 branches
(since there are 137,081 trees and 1,998 branches in each
tree) rather than 508,125 edges, resulting in $\sim$500 times more work.
% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
Even if we have a tree sequence consisting of a single tree
(negating the advantage of working edge-by-edge),
\msprime's mutation generator is still very efficient.
For example, we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim$260,000 mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods, because at this scale it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms.
Given these timings, it seems unlikely
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.

>>>>

We have greatly extended the sophistication of
\msprime's mutation generation engine for version 1.0,
achieving near feature-parity with \SeqGen.
We support a large number of mutation models, including the
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
Other models, such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation},
can be defined easily and efficiently in
user code by specifying a transition matrix between
any number of alleles.
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
\Pyvolve~\citep{spielman2015pyvolve}.

@@

Simulating mutations in \msprime\ is efficient.
Fig.~\ref{fig-mutations-perf} shows the time required to generate
mutations (using the default JC69 model) on
simulated tree sequences for a variety of mutation
rates as we vary the number of samples
(Fig.~\ref{fig-mutations-perf}A) and the sequence
length (Fig.~\ref{fig-mutations-perf}B).
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with 508,125 edges.
Simulating mutations on a single tree is also very efficient;
for example,
we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim$260,000 mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods, because at this scale it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms.
Given the above timings, it seems unlikely
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.
